---
title: "Reproducible Research"
output: 
  html_document:
    keep_md: true
---
  
    
```{r, include=FALSE}
library(data.table)
library(xtable)
library(ggplot2)
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

## Loading and preprocessing the data

```{r}
data_dir <- "C:/Users/RW/Documents";
fileURL <- "https://github.com/Arcapalo/RepData_PeerAssessment1/" 
fileSource <-"activity.zip"
source_path <- paste(data_dir, "/", fileSource , sep="")
txt_file <- paste(data_dir, "/","activity.csv", sep="")

if (!file.exists(txt_file)) {
        if (!file.exists(source_path)) {
                message(paste("Please Wait! Load", fileURL, "..."));
                download.file(fileURL, destfile=source_path);
        } 
        else {
            message(paste("Please Wait! Unzip", source_path, " file..."));
            unzip(source_path, exdir = data_dir);
        }
}
message(paste("Please Wait! Load", txt_file, " to dataset..."));
activity <- read.csv(txt_file, header=TRUE,   sep=",",
                    colClasses=c("numeric", "character", "numeric"))
activity$interval <- factor(activity$interval)
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
```
## What is mean total number of steps taken per day?

```{r, echo=FALSE}
steps_per_day <- aggregate(steps ~ date, activity, sum)
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "darkred", binwidth = 1000) + 
        labs(title="Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times (Count)") + 
        theme_bw()    
mean_steps = format(round(mean(steps_per_day$steps),2), scientific = FALSE)
median_steps = format(round(median(steps_per_day$steps),2), scientific = FALSE)

```

The MEAN of total number of steps taken per day is `r mean_steps`  
The MEDIAN of total number of steps taken per day is `r median_steps`

## What is the average daily activity pattern?

```{r, echo=FALSE}
steps_time_interval <- aggregate(activity$steps, 
                                by = list(interval = activity$interval),
                                FUN=mean, na.rm=TRUE)

steps_time_interval$interval <- as.integer(levels(steps_time_interval$interval)[steps_time_interval$interval])
colnames(steps_time_interval) <- c("interval", "steps")

ggplot(steps_time_interval, aes(x=interval, y=steps)) +   
        geom_line(color="blue", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw() + theme(legend.position = "bottom")
max_step_interval <- steps_time_interval[which.max(steps_time_interval$steps),]$interval
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
`r max_step_interval`  

## Inputing missing values

```{r, echo=FALSE}
missing <- nrow(activity)-nrow(na.omit(activity))
```

The total number of missing values in the dataset is `r missing` rows.

```{r}
## Filling in missing values  
fill_missing <- function(data, defaults) {
        na_indices <- which(is.na(data$steps))
        na_replacements <- unlist(lapply(na_indices, FUN=function(idx){
                interval = data[idx,]$interval
                defaults[defaults$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_indices] <- na_replacements
        fill_steps
}

##  New dataset with the missing data filled in
data_fill <- data.frame(  
        steps = fill_missing(activity, steps_time_interval),  
        date = activity$date,  
        interval = activity$interval)
```

```{r, echo=FALSE}
## Total number of steps taken each day histogram
full_steps_per_day <- aggregate(steps ~ date, data_fill, sum)
colnames(full_steps_per_day) <- c("date", "steps")
    
ggplot(full_steps_per_day, aes(x=steps)) + 
        geom_histogram(fill="darkred", binwidth=1000) + 
        labs(title="Full Steps Taken per Day", 
             x="Number of Steps with filled in missing values", 
             y="Count") + 
        theme_bw()     
        
full_mean_steps = format(round(mean(full_steps_per_day$steps), 2), scientific = FALSE)
full_median_steps = format(round(median(full_steps_per_day$steps), 2), scientific = FALSE)
```

Mean of total number of steps (with missing) taken per day is `r full_mean_steps`.  
Median of total number of steps (with missing) taken per day is `r full_median_steps`.  

These values DO differ from the estimates from the not filled ones 

The MEAN of total number of steps taken per day is `r mean_steps`  
The MEDIAN of total number of steps taken per day is `r median_steps`

Despite missing values had increased the peak, the impact on the predictions is not significant  


  
##Are there differences in activity patterns between weekdays and weekends?
```{r, echo=FALSE}
## Indicating whether a given date is a weekday or weekend day.
wkdays_steps <- function(data) {
    wkdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    wkdays_steps$interval <- 
            as.integer(levels(wkdays_steps$interval)[wkdays_steps$interval])
    colnames(wkdays_steps) <- c("interval", "steps")
    wkdays_steps
}

data_by_wkdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays in portuguese
    weekend_data <- subset(data, weekday %in% c("s bado","domingo"))
    weekday_data <- subset(data, !weekday %in% c("s bado","domingo"))
    
    weekend_steps <- wkdays_steps(weekend_data)
    weekday_steps <- wkdays_steps(weekday_data)
    
    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
    
    data_by_wkdays <- rbind(weekend_steps, weekday_steps)
    data_by_wkdays$dayofweek <- as.factor(data_by_wkdays$dayofweek)
    data_by_wkdays
}

data_weekdays <- data_by_wkdays(data_fill)
```

```{r, echo=FALSE}
## Average number of steps taken per 5-minute interval on weekdays and weekends
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="blue", size=1) + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```
